Gut microbes and their genes or something else

### 2023-02-20 manuscript/main.tex
get refs working

\addbibresource{zotero_lib}

% Highlighting
\usepackage{color}
\usepackage{soul}
\usepackage{xurl}
\usepackage{longtable,booktabs,array}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{assets}

@@

The gut and the brain are intimately linked. Signals from the brain
reach the gut through the autonomic nervous system and the endocrine
system, and the gut can communicate with the brain through the vagus
nerve and through endocrine and immune (cytokine) signaling molecules
\cite{cerdoEarlyNutritionGut2019,pronovostPerinatalInteractionsMicrobiome2019,sharonCentralNervousSystem2016,togniniGutMicrobiotaPotential2017}
In addition, the products of microbial metabolism generated in the gut can
influence the brain, both indirectly by stimulating the enteric nervous
and immune systems, as well as directly through molecules that enter
circulation and cross the blood-brain barrier. Causal links between the
gut microbiome and neural development, particularly atypical
development, are increasingly being identified
\cite{spichakMiningMicrobesMental2021}.
Both human epidemiology and animal models point to the effects
of gut microbes on the development of autism spectrum disorder
\cite{laueProspectiveAssociationsInfant2020,wanUnderdevelopmentGutMicrobiota2021}
and particular microbial taxa have been associated with depression
\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,valles-colomerNeuroactivePotentialHuman2019}
and Alzheimer's disease
\cite{fungInteractionsMicrobiotaImmune2017,kimProbioticSupplementationImproves2021}.
But information about this "microbiome-gut-brain
axis" in normal neurocognitive development remains lacking,
particularly early in life.

The first years of life are critical developmental windows for both the
microbiome and the brain
\cite{laueDevelopingMicrobiomeBirth2022}.
Fetal development \emph{in utero} is believed to be mostly sterile, but
is rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food sources (breastmilk or formula),
and other environmental sources such as antibiotics
\cite{backhedDynamicsStabilizationHuman2015,bokulichAntibioticsBirthMode2016}.
The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and domination by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\cite{koenigSuccessionMicrobialConsortia2011}.
Many of these microbes have specialized metabolic capabilities
for digesting human breast milk, such as \emph{Bifidobacterium infantis}
and \emph{Bacteroides fragilis}
\cite{selaGenomeSequenceBifidobacterium2008}.
Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation, with most taxa
of the infant microbiome being replaced by taxa more reminiscent of
adult microbiomes
\cite{backhedDynamicsStabilizationHuman2015}.
Many prior studies have focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
neural synaptogenesis, investigation across this solid-food boundary is
incredibly important
\cite{tauNormalDevelopmentBrain2010}.

A child's brain undergoes remarkable anatomical, microstructural,
organizational, and functional changes in the first years of life. By
age 5, a child's brain has reached \textgreater85\% of its adult size,
has achieved near-adult levels of myelination, and the pattern of axonal
connections has been established
\cite{silbereisCellularMolecularLandscapes2016}.
Much of this development occurs in discrete windows or
``sensitive-'' or ``critical periods'' (CPs)
\cite{knudsenSensitivePeriodsDevelopment2004}
when neural plasticity is particularly high,
and particular modes of learning
and skill development are preferred. The timing of these sensitive
periods is driven in part by genetics, but can also be affected by the
environment, including the gut microbiome
\cite{cowanAnnualResearchReview2020}.
In fact, emerging evidence suggests that the timing and duration of CPs
may be driven in part by cues from the developing gut microbiome
\cite{callaghanNestedSensitivePeriods2020}.
As such,
understanding the normal spectrum of healthy microbiome
development and how it relates to normal neurocognitive development may
provide opportunities for identifying atypical development earlier and
offer opportunities for intervention.

@@

We investigated the co-development of the brain and the microbiome in
early childhood in a cohort of typically-developing children in their
first years of life using a variety of orthogonal microbial and
neurocognitive assessments, including shotgun metagenomic sequencing,
cognitive and behavioral assessments, and neuroimaging. In all, 361
children from RESONANCE,
an accelerated longitudinal cohort of early development
\cite{forrestAdvancingScienceChildren2018}
between the ages of 82 days and 10 years
old (median age 2.21 years, Supplementary Figure 1) were included in
this study (Table 1, Figure 1A). To measure cognitive function, we used
age-appropriate assessments that can be normalized to a common, IQ-like
scale (Figure 1B), including the Mullen scale of early learning (MSEL)
for children from birth to 3 years of age
\cite{mullenMullenScalesEarly1995},
Wechsler preschool and primary scale of intelligence (WPPSI)
for 4-5 year-olds
\cite{wechslerWechslerPreschoolPrimary2012},
and the Wechsler intelligence scale for children (WISC)
for children 6 years and up
\cite{wechslerWechslerIntelligenceScale1949}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure1.png}
    \caption{
        Study design and overview. (\textbf{A}) Stool samples, cognitive assessments and
        neuroimaging were collected from participants in an accelerated
        longitudinal design. (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized to a common IQ-like scale.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }
    \label{fig:Figure 1}
\end{figure}

@@

Several studies have demonstrated links between specific taxa and
measures of anxiety and depression
\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022},
cognitive flexibility in adults
\cite{magnussonRelationshipsDietrelatedChanges2015},
and atypical neural development
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021}.
We, therefore, set out to identify whether specific taxa
or gene functions were linked with normal cognitive development in
children. In order to assess whether variations in gut microbial taxa,
their genes, or their metabolism are linked with neurocognitive
development, we tested whether the beta diversity of microbial taxa and
gene functions, as well as variation in brain development as assessed by
neuroimaging, were associated with these contemporaneous measures of
cognitive function using permutation analysis of variance (PERMANOVA).
Due to the large ecological shift in the microbiome that occurs upon the
introduction of solid food, and a relatively wide range of ages when
infants are transitioned to solid food, we considered the pre-transition
(less than 6 months old) and post-transition (over 18 months old)
microbiomes separately. Even still, age was a major driver of variation
in gut microbiomes in children over 18 months old (Figure 1G). We also
found that overall variation in microbial species in children over 18
months old was associated with a small but significant variation in
cognitive function score (Figure 1G, R\textsuperscript{2} = 0.0124, q
\textless{} 0.001), as was variation in microbial gene functions
(R\textsuperscript{2} = 0.0119, q \textless{} 0.001). Variation in
microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity, and broad lack of overlap between taxa in infants.
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and UniRef90s
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles (annotated with clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007}
), both generated from DNA sequencing data, being tightly coupled
(Figure 1H, p \textless{} 0.001). Interestingly, other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and
KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004}
) overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (15.9\%, p \textless{} 0.01) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression
\cite{mallickMultivariableAssociationDiscovery2021}
to the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N = 92 for 0--120 months, N
= 46 for 0--6 months, N = 97 for 18--120 months). Only \emph{Blautia wexlerae}
was significantly associated (q value = 0.14, $\beta$ = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B). \emph{B. wexlerae} was
previously shown to be depleted in children with diabetes
\cite{benitez-paezDepletionBlautiaSpecies2020},
and that oral administration of \emph{B. wexlerae} partially
ameliorated weight gain and inflammation from a high-fat diet in a mouse
model of T2D
\cite{hosomiOralAdministrationBlautia2022,liuBlautiaNewFunctional2021}.
In children over 18 months of age,
several microbial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae}, which produces the
neuroprotective metabolite urolithin
\cite{gongUrolithinAlleviatesBloodbrain2022,selmaDescriptionUrolithinProduction2014},
\emph{Asaccharobacter celatus} and
\emph{Adelcreutzia equolifaciens}, which produce phytoestrogen-derived
equol
\cite{maruoAdlercreutziaEquolifaciensGen2008,thawornkunoBiotransformationDaidzeinEquol2009},
and the SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B).

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        \textbf{Figure 2 - Taxa and gene functional groups are associated with cognitive function.}
        (A) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age, controlling for age, maternal education,
        and sequencing depth. Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (B) For taxa that were significantly
        associated with cognitive function, heatmaps of prevalence, mean
        relative abundance, and correlation with cognitive function in different
        age groups. (C-F) Enrichment plots for selected neuroactive gene sets
        used in feature set enrichment analysis (FSEA). (G) summary of FSEA
        results across all samples (left) as well as in the under 6 months
        (middle) and over 18 months (right) subsets, colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set, and vertical bars represent the
        median correlation of that gene set.
    }
    \label{fig:Figure 2}
\end{figure}

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive function score (Table 2, Figure 2C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive function scores. This was true both when
considering all age groups together, though the enrichment of most
pathways was more pronounced in children under 6 months or over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive function
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function \cite{dalileRoleShortchainFatty2019}.

Synthesis of menaquinone (vitamin K) was also negatively associated with
cognitive function score in older children (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183). Menaquinone has several isoforms, one of
which, MK-4, has been found to be decreased in children diagnosed with ASD
\cite{dongCorrelationSerumConcentrations2021}
and is potentially neuroprotective in both rodents and humans
\cite{elkattawyVitaminK2Menaquinone72022}.
In children over 18 months old,
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (Glutamate synthesis I, E.S. = 0.242, q
= 0.047; Tryptophan synthesis, E.S. = 0.119, q = 0.041). Glutamate is a
critical neurotransmitter controlling neuronal excitatory/inhibitory
signaling along with gamma-aminobutyric acid (GABA), and their balance
in the brain controls neural plasticity and learning, particularly in
the developing brain
\cite{cohenkadoshLinkingGABAGlutamate2015,palomo-buitragoGlutamateInteractionsObesity2019}.
Tryptophan metabolism, including
microbial metabolism of tryptophan, has previously been linked with
autism in children
\cite{hoshinoBloodSerotoninFree1984,xiaoFecalMicrobiomeTransplantation2021}.
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

@@

FSEA relies on understanding functional relationships between individual
genes. However, because the relationships between individual taxa are
still largely unknown, we turned to Random Forest (RF) models, an
unsupervised non-parametric machine learning (ML) method that enables
the identification of underlying patterns in large numbers of individual
features (here, microbial species). Previous studies have reported
successful use of RFs for processing highly-dimensional and sparse data
from the domain of genomics
\cite{amaratungaEnrichedRandomForests2008,brieucPracticalIntroductionRandom2018,chenRandomForestsGenomic2012,franzosaGutMicrobiomeStructure2019,stephanRandomForestApproach2015},
along with other works where it was used
to predict cognitive conditions related to Alzheimer's disease in
different scenarios
\cite{ardekaniPredictionIncipientAlzheimer2017,velazquezRandomForestModel2021}.
Additionally, given the sequential
nature of variable consideration in each tree, RFs are naturally able to
work out complex input feature interactions, such as those present in a
microbiome-wide study, without the necessity to explicitly compute
interaction terms.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidt]{assets/Figure3.png}
    \caption{
        \textbf{Figure 3 - Random forest models predict concurrently measured cognitive function.}
        Comparison of RF predictor importance versus linear models for children
        between birth and 6 months old (A), and for those older than 18 months
        (B). Colors represent whether the species belong to the group of
        top-important features that account for 80\% of the cumulative
        importance on the RF model, if that species was significant (q
        \textless{} 0.2) in linear models, both, or neither. Ranked feature
        importance for taxa in RF models for children between birth and 6 months
        old (C), and for those older than 18 months (D). Taxa that are important
        for RF models both for children under 6 months and for those over 18
        months (E), only for children under 6 months (F), and only for children
        over 18 months (G)
    }
    \label{fig:Figure 3}
\end{figure}

@@

We also observed that many brain regions had high accordance across the
left and right hemispheres in terms of both model performance and
microbial feature importance. In contrast, other regions had substantial
differences between the hemispheres. For example, the left accumbens
area, which plays an important role in reward circuits
\cite{ernstAmygdalaNucleusAccumbens2005,yauNucleusAccumbensResponse2012},
has one of the highest test-set correlations of our
brain region models (R = 0.288), as compared to the right accumbens
models which could not adequately generalize, and had a negative mean
test-set correlation (R = -0.041). Most healthy individuals have a
rightward asymmetry in the nucleus accumbens, and reduced asymmetry has
been linked to substance use disorder in young adults
\cite{caoMappingCorticalSubcortical2021}.

Feature importance for models of the left accumbens area were dominated
by three species of \emph{Bacteroides}, \emph{B. vulgatus} (3.4\%
relative importance), \emph{B. ovatus} (3.7\% relative importance), and
\emph{B. uniformis} (3.0\% relative importance).. The accumbens area is
associated with reward control, and in individuals diagnosed with ADHD,
it has been shown to have a divergent neuromorphology
\cite{hoogmanSubcorticalBrainVolume2017}.
Independently, \emph{B. ovatus}, \emph{B. uniformis} and
\emph{B. vulgatus} have been linked to ADHD
\cite{wangGutMicrobiotaDietary2020}.

>>>>

\addbibresource{zotero.bib}

% Highlighting
\usepackage{color}
\usepackage{soul}
\usepackage{xurl}
\usepackage{longtable,booktabs,array}
\usepackage{multirow}

% figures
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{graphicx}
\graphicspath{assets}

@@

The gut and the brain are intimately linked. Signals from the brain
reach the gut through the autonomic nervous system and the endocrine
system, and the gut can communicate with the brain through the vagus
nerve and through endocrine and immune (cytokine) signaling molecules
\citep{cerdoEarlyNutritionGut2019,pronovostPerinatalInteractionsMicrobiome2019,sharonCentralNervousSystem2016,togniniGutMicrobiotaPotential2017}
In addition, the products of microbial metabolism generated in the gut can
influence the brain, both indirectly by stimulating the enteric nervous
and immune systems, as well as directly through molecules that enter
circulation and cross the blood-brain barrier. Causal links between the
gut microbiome and neural development, particularly atypical
development, are increasingly being identified
\citep{spichakMiningMicrobesMental2021}.
Both human epidemiology and animal models point to the effects
of gut microbes on the development of autism spectrum disorder
\citep{laueProspectiveAssociationsInfant2020,wanUnderdevelopmentGutMicrobiota2021}
and particular microbial taxa have been associated with depression
\citep{mayneris-perxachsMicrobiotaAlterationsProline2022,valles-colomerNeuroactivePotentialHuman2019}
and Alzheimer's disease
\citep{fungInteractionsMicrobiotaImmune2017,kimProbioticSupplementationImproves2021}.
But information about this "microbiome-gut-brain
axis" in normal neurocognitive development remains lacking,
particularly early in life.

The first years of life are critical developmental windows for both the
microbiome and the brain
\citep{laueDevelopingMicrobiomeBirth2022}.
Fetal development \emph{in utero} is believed to be mostly sterile, but
is rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food sources (breastmilk or formula),
and other environmental sources such as antibiotics
\citep{backhedDynamicsStabilizationHuman2015,bokulichAntibioticsBirthMode2016}.
The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and domination by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\citep{koenigSuccessionMicrobialConsortia2011}.
Many of these microbes have specialized metabolic capabilities
for digesting human breast milk, such as \emph{Bifidobacterium infantis}
and \emph{Bacteroides fragilis}
\citep{selaGenomeSequenceBifidobacterium2008}.
Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation, with most taxa
of the infant microbiome being replaced by taxa more reminiscent of
adult microbiomes
\citep{backhedDynamicsStabilizationHuman2015}.
Many prior studies have focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
neural synaptogenesis, investigation across this solid-food boundary is
incredibly important
\citep{tauNormalDevelopmentBrain2010}.

A child's brain undergoes remarkable anatomical, microstructural,
organizational, and functional changes in the first years of life. By
age 5, a child's brain has reached \textgreater85\% of its adult size,
has achieved near-adult levels of myelination, and the pattern of axonal
connections has been established
\citep{silbereisCellularMolecularLandscapes2016}.
Much of this development occurs in discrete windows or
``sensitive-'' or ``critical periods'' (CPs)
\citep{knudsenSensitivePeriodsDevelopment2004}
when neural plasticity is particularly high,
and particular modes of learning
and skill development are preferred. The timing of these sensitive
periods is driven in part by genetics, but can also be affected by the
environment, including the gut microbiome
\citep{cowanAnnualResearchReview2020}.
In fact, emerging evidence suggests that the timing and duration of CPs
may be driven in part by cues from the developing gut microbiome
\citep{callaghanNestedSensitivePeriods2020}.
As such,
understanding the normal spectrum of healthy microbiome
development and how it relates to normal neurocognitive development may
provide opportunities for identifying atypical development earlier and
offer opportunities for intervention.

@@

We investigated the co-development of the brain and the microbiome in
early childhood in a cohort of typically-developing children in their
first years of life using a variety of orthogonal microbial and
neurocognitive assessments, including shotgun metagenomic sequencing,
cognitive and behavioral assessments, and neuroimaging. In all, 361
children from RESONANCE,
an accelerated longitudinal cohort of early development
\citepp{forrestAdvancingScienceChildren2018}
between the ages of 82 days and 10 years
old (median age 2.21 years, Supplementary Figure 1) were included in
this study (Table 1, Figure 1A). To measure cognitive function, we used
age-appropriate assessments that can be normalized to a common, IQ-like
scale (Figure 1B), including the Mullen scale of early learning (MSEL)
for children from birth to 3 years of age
\citep{mullenMullenScalesEarly1995},
Wechsler preschool and primary scale of intelligence (WPPSI)
for 4-5 year-olds
\citep{wechslerWechslerPreschoolPrimary2012},
and the Wechsler intelligence scale for children (WISC)
for children 6 years and up
\citep{wechslerWechslerIntelligenceScale1949}.

@@

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure1.png}
    \caption{
        Study design and overview. (\textbf{A}) Stool samples, cognitive assessments and
        neuroimaging were collected from participants in an accelerated
        longitudinal design. (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized to a common IQ-like scale.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }
    \label{fig:Figure 1}
\end{sidewaysfigure}

@@

Several studies have demonstrated links between specific taxa and
measures of anxiety and depression
\citep{mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022},
cognitive flexibility in adults
\citep{magnussonRelationshipsDietrelatedChanges2015},
and atypical neural development
\citep{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021}.
We, therefore, set out to identify whether specific taxa
or gene functions were linked with normal cognitive development in
children. In order to assess whether variations in gut microbial taxa,
their genes, or their metabolism are linked with neurocognitive
development, we tested whether the beta diversity of microbial taxa and
gene functions, as well as variation in brain development as assessed by
neuroimaging, were associated with these contemporaneous measures of
cognitive function using permutation analysis of variance (PERMANOVA).
Due to the large ecological shift in the microbiome that occurs upon the
introduction of solid food, and a relatively wide range of ages when
infants are transitioned to solid food, we considered the pre-transition
(less than 6 months old) and post-transition (over 18 months old)
microbiomes separately. Even still, age was a major driver of variation
in gut microbiomes in children over 18 months old (Figure 1G). We also
found that overall variation in microbial species in children over 18
months old was associated with a small but significant variation in
cognitive function score (Figure 1G, R\textsuperscript{2} = 0.0124, q
\textless{} 0.001), as was variation in microbial gene functions
(R\textsuperscript{2} = 0.0119, q \textless{} 0.001). Variation in
microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity, and broad lack of overlap between taxa in infants.
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and UniRef90s
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles (annotated with clusters of 90\% similarity
\citep{suzekUniRefComprehensiveNonredundant2007}
), both generated from DNA sequencing data, being tightly coupled
(Figure 1H, p \textless{} 0.001). Interestingly, other functional
groupings (Enzyme commission level4 - ECs
\citep{bairochENZYMEDatabase20002000},
and
KEGG orthologues - KOs
\citep{kanehisaKEGGResourceDeciphering2004}
) overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (15.9\%, p \textless{} 0.01) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression
\citep{mallickMultivariableAssociationDiscovery2021}
to the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N = 92 for 0--120 months, N
= 46 for 0--6 months, N = 97 for 18--120 months). Only \emph{Blautia wexlerae}
was significantly associated (q value = 0.14, $\beta$ = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B). \emph{B. wexlerae} was
previously shown to be depleted in children with diabetes
\citep{benitez-paezDepletionBlautiaSpecies2020},
and that oral administration of \emph{B. wexlerae} partially
ameliorated weight gain and inflammation from a high-fat diet in a mouse
model of T2D
\citep{hosomiOralAdministrationBlautia2022,liuBlautiaNewFunctional2021}.
In children over 18 months of age,
several microbial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae}, which produces the
neuroprotective metabolite urolithin
\citep{gongUrolithinAlleviatesBloodbrain2022,selmaDescriptionUrolithinProduction2014},
\emph{Asaccharobacter celatus} and
\emph{Adelcreutzia equolifaciens}, which produce phytoestrogen-derived
equol
\citep{maruoAdlercreutziaEquolifaciensGen2008,thawornkunoBiotransformationDaidzeinEquol2009},
and the SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\citep{ghoshMediterraneanDietIntervention2020}
(Figure 2B).

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        Taxa and gene functional groups are associated with cognitive function.
        (A) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age, controlling for age, maternal education,
        and sequencing depth. Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (B) For taxa that were significantly
        associated with cognitive function, heatmaps of prevalence, mean
        relative abundance, and correlation with cognitive function in different
        age groups. (C-F) Enrichment plots for selected neuroactive gene sets
        used in feature set enrichment analysis (FSEA). (G) summary of FSEA
        results across all samples (left) as well as in the under 6 months
        (middle) and over 18 months (right) subsets, colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set, and vertical bars represent the
        median correlation of that gene set.
    }
    \label{fig:Figure 2}
\end{sidewaysfigure}

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\citep{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive function score (Table 2, Figure 2C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive function scores. This was true both when
considering all age groups together, though the enrichment of most
pathways was more pronounced in children under 6 months or over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive function
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function \citep{dalileRoleShortchainFatty2019}.

Synthesis of menaquinone (vitamin K) was also negatively associated with
cognitive function score in older children (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183). Menaquinone has several isoforms, one of
which, MK-4, has been found to be decreased in children diagnosed with ASD
\citep{dongCorrelationSerumConcentrations2021}
and is potentially neuroprotective in both rodents and humans
\citep{elkattawyVitaminK2Menaquinone72022}.
In children over 18 months old,
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (Glutamate synthesis I, E.S. = 0.242, q
= 0.047; Tryptophan synthesis, E.S. = 0.119, q = 0.041). Glutamate is a
critical neurotransmitter controlling neuronal excitatory/inhibitory
signaling along with gamma-aminobutyric acid (GABA), and their balance
in the brain controls neural plasticity and learning, particularly in
the developing brain
\citep{cohenkadoshLinkingGABAGlutamate2015,palomo-buitragoGlutamateInteractionsObesity2019}.
Tryptophan metabolism, including
microbial metabolism of tryptophan, has previously been linked with
autism in children
\citep{hoshinoBloodSerotoninFree1984,xiaoFecalMicrobiomeTransplantation2021}.
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

@@

FSEA relies on understanding functional relationships between individual
genes. However, because the relationships between individual taxa are
still largely unknown, we turned to Random Forest (RF) models, an
unsupervised non-parametric machine learning (ML) method that enables
the identification of underlying patterns in large numbers of individual
features (here, microbial species). Previous studies have reported
successful use of RFs for processing highly-dimensional and sparse data
from the domain of genomics
\citep{amaratungaEnrichedRandomForests2008,brieucPracticalIntroductionRandom2018,chenRandomForestsGenomic2012,franzosaGutMicrobiomeStructure2019,stephanRandomForestApproach2015},
along with other works where it was used
to predict cognitive conditions related to Alzheimer's disease in
different scenarios
\citep{ardekaniPredictionIncipientAlzheimer2017,velazquezRandomForestModel2021}.
Additionally, given the sequential
nature of variable consideration in each tree, RFs are naturally able to
work out complex input feature interactions, such as those present in a
microbiome-wide study, without the necessity to explicitly compute
interaction terms.

\begin{sidewaysfigure}[h]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure3.png}
    \caption{
        Random forest models predict concurrently measured cognitive function.
        Comparison of RF predictor importance versus linear models for children
        between birth and 6 months old (A), and for those older than 18 months
        (B). Colors represent whether the species belong to the group of
        top-important features that account for 80\% of the cumulative
        importance on the RF model, if that species was significant (q
        \textless{} 0.2) in linear models, both, or neither. Ranked feature
        importance for taxa in RF models for children between birth and 6 months
        old (C), and for those older than 18 months (D). Taxa that are important
        for RF models both for children under 6 months and for those over 18
        months (E), only for children under 6 months (F), and only for children
        over 18 months (G)
    }
    \label{fig:Figure 3}
\end{sidewaysfigure}

@@

We also observed that many brain regions had high accordance across the
left and right hemispheres in terms of both model performance and
microbial feature importance. In contrast, other regions had substantial
differences between the hemispheres. For example, the left accumbens
area, which plays an important role in reward circuits
\citep{ernstAmygdalaNucleusAccumbens2005,yauNucleusAccumbensResponse2012},
has one of the highest test-set correlations of our
brain region models (R = 0.288), as compared to the right accumbens
models which could not adequately generalize, and had a negative mean
test-set correlation (R = -0.041). Most healthy individuals have a
rightward asymmetry in the nucleus accumbens, and reduced asymmetry has
been linked to substance use disorder in young adults
\citep{caoMappingCorticalSubcortical2021}.

Feature importance for models of the left accumbens area were dominated
by three species of \emph{Bacteroides}, \emph{B. vulgatus} (3.4\%
relative importance), \emph{B. ovatus} (3.7\% relative importance), and
\emph{B. uniformis} (3.0\% relative importance).. The accumbens area is
associated with reward control, and in individuals diagnosed with ADHD,
it has been shown to have a divergent neuromorphology
\citep{hoogmanSubcorticalBrainVolume2017}.
Independently, \emph{B. ovatus}, \emph{B. uniformis} and
\emph{B. vulgatus} have been linked to ADHD
\citep{wangGutMicrobiotaDietary2020}.
